\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=2cm}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{setspace}  % For line spacing control
\usepackage{tocloft}   % For Table of Contents formatting

% Set 1.5 line spacing for main text
\onehalfspacing

% Set font to Times New Roman (if available)
\usepackage{times}

% Set up word count command using texcount (requires -shell-escape)
\newcommand\wordcount{\input{|"texcount -1 -sum 'TE Classification.tex'"}}

\begin{document}

% ============ TITLE PAGE ============
\begin{titlepage}
    \centering
    \vspace*{2cm}
    
    % Title
    {\LARGE\bfseries Machine Learning a Transposable Element Classifier\par}
    
    \vspace{3cm}
    
    % Author name
    {\Large Yingrui Yang\par}
    
    \vspace{3cm}
    
    % Supervisor information
    {\large 
    Supervisor: Richard Durbin\par
    \vspace{0.5cm}
    Day-to-day Supervisors: Pio Sierra, Hadi Khan\par
    }
    
    \vfill
    
    % Word count statement
    {\normalsize This thesis contains approximately \wordcount\ words.\par}
    
    \vspace{1cm}
    
\end{titlepage}

% ============ DECLARATION PAGE ============
\newpage
\thispagestyle{empty}

\vspace*{1cm}

\begin{center}
    {\Large\bfseries Declaration\par}
\end{center}

\vspace{1cm}

\noindent\textbf{Name:} Yingrui Yang

\vspace{0.5cm}

\noindent\textbf{College:} Emmanuel College

\vspace{0.5cm}

\noindent\textbf{Title:} Machine Learning a Transposable Element Classifier

\vspace{1.5cm}

\noindent I understand the University's definition of plagiarism. I declare that, in accordance with Discipline regulation 6, this dissertation is entirely my own work except where otherwise stated, either in the form of citation of published work, or acknowledgement of the source of any unpublished material.

\vspace{1.5cm}

\noindent The project is purely computational and does not involve any wet lab experiments. The models were trained on CSD3 in the Western Cambridge Data Center, with the access provided by the research group. 

\vspace{2cm}

\noindent\textbf{Signature:} \hrulefill

\vspace{1cm}

\noindent\textbf{Date:} \hrulefill

% ============ BLIND GRADING NUMBER PAGE ============
\newpage
\thispagestyle{empty}

\vspace*{\fill}
\begin{center}
    {\fontsize{24.88}{34}\selectfont\bfseries Part III project report 6506D\par}
\end{center}
\vspace*{\fill}

% ============ SUMMARY ============
\newpage
\section*{Summary}
\addcontentsline{toc}{section}{Summary}

[PLACEHOLDER: Summary]

% ============ TABLE OF CONTENTS ============
\newpage
\tableofcontents

% ============ LIST OF ABBREVIATIONS ============
\newpage
\section*{List of Abbreviations}
\addcontentsline{toc}{section}{List of Abbreviations}

\begin{tabular}{ll}
    TE & Transposable Element \\
    CNN & Convolutional Neural Network \\
    GNN & Graph Neural Network \\
\end{tabular}

\iffalse

%% Supervisor's plan for project - to be used as a basis for writing the thesis, but not to be included in the final document

Transposable elements (TEs) are kilobase-scale segments of DNA that insert new copies of themselves in the genome, generating over time large families of interspersed repeats (ref 1). Some of these are autonomous, themselves encoding the proteins needed to mobilise themselves, but others are non-autonomous, needing an active autonomous element to move. Our group with Prof. Karam Teixeira’s is studying the co-evolution of TE families and subfamilies with eukaryote genomes. Although we have increasingly good methods to identify TE families in new genome sequences, classifying them correctly is not a solved problem. In particular it is frequently difficult to correctly identify non-autonomous element subtypes and assign them to their driver elements. Current methods rely on detection of a set of features, with manual curation, but recently machine learning methods have been explored. In this project the student will take advantage of very large sets of TEs found using our new detection approach Pantera (ref 2), to build better classifiers using novel machine learning approaches, for example using DNA and protein foundation models and unsupervised learning to identify new families, and exploring new approaches to associate non-autonomous sequences to their drivers. One goal is to build generative models of TE sequences that encode information not captured by classic sequence alignment.

1. Wells JN, Feschotte C (2020) A Field Guide to Eukaryotic Transposable Elements. Annu Rev Genet. 54:539-561. https://doi.org/10.1146/annurev-genet-040620-022145 

2. Sierra P, Durbin R (2024) Identification of transposable element families from pangenome polymorphisms. Mobile DNA. 15:13. https://doi.org/10.1186/s13100-024-00323-y 

3. Balamurugan R, Mohite S, Raja SP (2023) Protein Sequence Classification Using Bidirectional Encoder Representations from Transformers (BERT) Approach. SN COMPUT. SCI. 4:481. https://doi.org/10.1007/s42979-023-01980-1 

4. David KT, Halanych KM (2023) Unsupervised Deep Learning Can Identify Protein Functional Groups from Unaligned Sequences, Genome Biology and Evolution, 15:evad084, https://doi.org/10.1093/gbe/evad084 

\fi

\iffalse

%% This is NOT the interium report. I could not find the original document, but this is a very relevant email and contains some high-level summaries of the work done by the time. Again, just for reference and not to be included in the final document.

CNN Tower Model (v3)

This is an update to the models I trained before the vacation, with the tower-based CNN network and RC-invariance via fusion. The model generally goes through three CNN layers with kernel sizes 7, 15, 21 to capture difference motifs, and the results are put into attention layer. The model currently has a binary head to deal with transposase vs None and a multi classification head (among 7 families since I'm ruling out all the families with fewer then 100 samples). 

I'm getting pretty decent results, with 91.3% accuracy and 0.87 Macro F1 on the binary head, and 94.4% accuracy and 0.90 Macro F1 on the multi head. Please see the confusion matrix as attached. The model performs especially well on DNA/hAT (F1=0.98) and TcMar-Tc1 (F1=0.97), while DNA/CMC remains challenging (F1=0.76), likely due to limited training samples.

I'm placing the raw sequences at random positions within a 40,000 bp window during training. Interestingly, the model learned to predict sequence length with high accuracy (correlation 0.94, MAE ~1kb), though absolute boundary positions remain difficult to localise.

CNN Tower + k-mer GNN Hybrid Model (v4)

As previously discussed, I experimented with combining a windowed k-mer GNN to help the model learn compositional patterns across the sequence. The standalone k-mer GNN does not perform as well on its own as it tends to confuse superfamilies within the DNA class since they share similar k-mer compositions. However, it provides useful complementary information, so I combined both approaches using attention-based fusion.

In the best trained epoch, the attention fusion weights settled at approximately CNN: 40% / GNN: 60%, but this proportion is quite stable at the late stage of training. The hybrid model doesn't outperform v3 on superfamily classification but is slightly better at the binary head. 

Model using the full label set (v5, in progress)

The previous models were trained with Pio's 2025-12-15 feature set, which labels over 92% sequences as "None" (no transposase detected). However, the FASTA headers themselves contain more detailed class labels (e.g., LTR/Gypsy, LINE/L1). I'm now training a model on this fuller label set, which may improve the binary classification by leveraging the richer supervision signal. 

Interim Report

Finally, regarding the interim report, I've just realised the deadline is Jan 15 and that supervisor and student need to submit separate reports. I apologise for the late notice on this. I'll be back in Cambridge on Tuesday (Jan 13), so I hope there will be enough time to complete it. Please let me know if you have any guidance or templates I should follow.

\fi

\iffalse

%% This is another quote from email. The first feature set for TIR is the starting point of the project, and the second feature set includes the classification labels for only the DNA elements, while everything else are labeled as "None". It is important to clarify that in the model files I was writing "Transposase vs None" for the second model, but it is not exactly correct since the "None" class includes all non-DNA elements, not just the ones without transposase. 

That looks good, but I am a little confused by the “Transposase vs None” model. Do you mean the first feature table? If so that is about presence/absence of TIR elements, the inverted repeats at the edges of the sequence.

If instead you mean “anything not none vs none” in the second table, that would put together all elements classified as DNA elements, but not all of them would have a transposase, so that is not what the model would be detecting.

\fi

% ============ INTRODUCTION ============
\newpage
\section{Introduction}

[PLACEHOLDER: Introduction]

% ============ RESULTS ============
\newpage
\section{Results}

\subsection{Terminal Inverted Repeat Detection}

Terminal inverted repeats (TIRs) are a defining structural feature of DNA transposons, consisting of near-identical sequences in opposite orientations at the element boundaries. Detecting TIR presence is a fundamental task for TE classification, as it distinguishes DNA transposons from other TE classes such as LTR retrotransposons and LINEs. I developed a convolutional neural network (CNN) specifically designed for binary TIR detection, addressing two key challenges: the inherent reverse-complement symmetry of TIRs and severe class imbalance in the training data.

\subsubsection{Model Architecture and Reverse-Complement Invariance}

A sequence containing a TIR should be recognized regardless of which strand is provided as input. In other words, the model should be invariant to the reverse-complement transformation for any sequence. To achieve this, I implemented a tower shaped CNN architecture with

The model architecture employs multi-scale motif detection with convolutional kernels of sizes 7, 15, and 21 base pairs, capturing both short sequence motifs and longer structural features characteristic of TIRs. These are followed by dilated convolutional blocks with dilation rates of 1, 2, 4, and 8, providing an exponentially increasing receptive field to capture long-range dependencies across the 40 kb input window.

\subsubsection{Handling Class Imbalance}

The training dataset exhibited substantial class imbalance, with TIR-positive sequences representing only approximately 8\% of the total data. I explored several approaches to address this imbalance, including inverse frequency class weighting, inverse square-root weighting, focal loss, and stratified subsampling. Surprisingly, the last approach - simply selecting a balanced subset of the data for training - outperformed more sophisticated loss-based approaches.

This finding has practical implications: rather than engineering complex loss functions that can lead to training instability, straightforward data balancing through subsampling provides robust and reproducible results. The model trained with subsampling achieved [XX]\% accuracy and [XX] F1 score on the held-out test set, with balanced performance across both classes.

\subsubsection{Sequence Position Augmentation}

To ensure the model learns position-invariant features, I placed each input sequence at a random position within the 40 kb canvas during training. This augmentation strategy forces the model to recognize TIR signatures regardless of their absolute position, improving generalization to sequences of varying lengths and genomic contexts.


\subsection{Hierarchical DNA Transposon Classification (v3)}\label{ssec:hier}

The core goal of the project is to build a classifier that can assign each transposable element to its correct superfamily based on sequence features, with particular focus on DNA transposons. I developed a hierarchical classification system that performs two tasks simultaneously: (1) binary classification of whether a sequence is a DNA transposon or belongs to another TE class (LTR retrotransposons, LINEs, SINEs, etc.), and (2) multi-class classification into superfamilies for DNA-positive sequences. This hierarchical approach reflects the biological organization of TEs and allows the model to leverage shared features across related classification tasks.

It is important to note that the binary task distinguishes DNA transposons from other TE classes, not the presence of transposase domains specifically. While most autonomous DNA transposons encode transposases, many sequences in the DNA class are non-autonomous elements (e.g., MITEs) that lack transposase but share structural features with their autonomous relatives.

\subsubsection{Hierarchical Design}

The model employs two classification heads branching from a shared CNN embedding: a binary head for DNA vs non-DNA classification and a multi-class head for superfamily assignment among DNA elements. This hierarchical design reflects the biological organization---DNA transposons share structural features (such as TIRs) that distinguish them from retrotransposons, while superfamily-specific sequence motifs provide finer discrimination. Training both heads jointly allows the model to learn representations useful for both tasks, with the binary head providing a strong gradient signal even for rare superfamilies.

\subsubsection{Addressing Extreme Class Imbalance}

The dataset presents a challenging imbalance: over 92\% of sequences are non-DNA elements (labeled ``None'' in the feature annotations), while the remaining 8\% are DNA transposons distributed across seven superfamilies with more than 100 samples each. I addressed this through strategic subsampling, reducing the ``None'' class from approximately 135,000 to 20,000 sequences while retaining all DNA transposon samples. This approach maintains sufficient negative examples for robust binary classification while preventing the model from defaulting to majority-class predictions.

Within the transposase-positive class, superfamily distribution remains imbalanced. Inverse square-root class weighting in the loss function upweights rare superfamilies without the instability that inverse frequency weighting can cause.

\subsubsection{Performance}

The hierarchical model achieves strong performance on both tasks. For binary classification (DNA transposon vs other TE classes), the model reaches 91.3\% accuracy with a macro F1 score of 0.87. For superfamily classification among DNA transposons, accuracy improves to 94.4\% with a macro F1 of 0.90.

Performance varies across superfamilies, correlating with both sample size and biological distinctiveness. DNA/hAT achieves the highest F1 score of 0.98, reflecting both its abundance in the training data and its distinctive sequence features. TcMar-Tc1 similarly performs well (F1 = 0.97). In contrast, DNA/CMC remains challenging (F1 = 0.76), likely due to limited training samples and greater sequence diversity within this superfamily.

The confusion matrix reveals that most misclassifications occur between biologically related superfamilies, suggesting the model captures meaningful evolutionary relationships rather than arbitrary patterns.


\subsection{Hybrid CNN and K-mer Graph Neural Network Model (v4)}\label{ssec:hybrid}

While the CNN-based hierarchical model performs well, convolutional architectures may not fully capture compositional patterns—the overall distribution of sequence motifs—that distinguish TE superfamilies. I hypothesized that combining local sequence features from CNNs with global compositional information could improve classification. To test this, I developed a hybrid architecture that fuses a CNN tower with a k-mer-based graph neural network (GNN).

\subsubsection{K-mer Graph Representation}

The GNN branch represents each sequence as a graph where nodes correspond to sliding windows and edges connect adjacent windows. Each node contains the k-mer frequency profile of its window, using canonical k-mer representation for RC-invariance. This windowed approach preserves positional information that would be lost with global k-mer counts, while the graph structure allows the GNN to propagate compositional information across the sequence.

\subsubsection{Attention-Based Fusion}

Rather than simply concatenating CNN and GNN embeddings, I employ learned attention-based fusion. A gating mechanism computes weights for each modality based on the input, allowing the model to dynamically emphasize the more informative branch for each sequence. The fused representation is passed to the same hierarchical classification heads as the CNN-only model.

In the binary classification setting, the learned gate weights stabilize at approximately 40\% CNN and 60\% GNN across training, suggesting that global compositional patterns provide stronger signal than local motifs for the broad DNA/non-DNA task. The ratio shifts when the model is extended to the three-class setting (Section~\ref{ssec:fullset}), where CNN becomes dominant.

\subsubsection{Comparison with CNN-Only Model}

The hybrid model shows modest improvements over the CNN-only architecture, particularly for the binary classification task (DNA vs non-DNA). However, superfamily classification performance is comparable between the two approaches. This suggests that while k-mer composition provides useful signal for distinguishing DNA transposons from other TE classes, the fine-grained distinctions between superfamilies are better captured by local sequence motifs.

The hybrid architecture also incurs computational overhead: k-mer feature extraction and GNN processing approximately double the training time per epoch. Given the marginal performance gains, the pure CNN model may be preferable for practical applications where computational efficiency is important.


\subsection{Multi-Class Extension: Three-Class Classification (v4.2 and v4.3)}\label{ssec:fullset}

The binary models described in Sections~\ref{ssec:hier} and~\ref{ssec:hybrid} classify DNA transposons against all other TEs lumped into a single ``None'' category. This formulation discards supervision available in the FASTA annotations, which assign sequences to specific classes reflecting distinct transposition mechanisms. I extended the hybrid CNN+GNN model to a three-class head distinguishing DNA transposons, LTR retrotransposons, and LINEs, developing two successive versions: v4.2, trained on the full VGP dataset, and v4.3, which reserves three benchmark species for external validation.

\subsubsection{Three-Class Reformulation}

The three-class head asks the model to distinguish: cut-and-paste DNA transposons; LTR retrotransposons, which copy via an RNA intermediate flanked by long terminal repeats; and LINEs, which replicate through target-primed reverse transcription. Sequences with ambiguous or rare class labels (None, SINE, PLE, RC) are excluded from training and evaluation, leaving 50,431 sequences distributed across 23 superfamilies. The CNN and GNN tower architectures are unchanged from the hybrid model (Section~\ref{ssec:hybrid}); only the top-level classification head is broadened from two outputs to three, with superfamily subheads spanning all three major classes.

\subsubsection{Overfitting Analysis in v4.2}

The first three-class version (v4.2) was trained on the full VGP dataset for up to 65 epochs. Although the checkpoint-selection metric (validation macro F1) appeared saturated---with both training and validation superfamily F1 reaching 0.999---the held-out test set revealed a substantial generalization gap: superfamily macro F1 dropped to 0.872 on the test set, a gap of 0.127 (moderate overfitting). Top-level class F1 generalized markedly better (test F1 = 0.984, gap = 0.016), indicating that three-class discrimination is inherently robust while fine-grained superfamily classification is susceptible to memorizing genome-specific sequence patterns. This analysis prompted two refinements carried into v4.3: capping training at 40 epochs and withholding three species as a clean external benchmark.

\subsubsection{K-mer Features Separate Top-Level Classes}

Analysis of k-mer features reveals that compositional patterns alone largely distinguish the three major TE classes. Logistic regression and random forest classifiers trained on 7-mer frequencies achieve approximately 85\% macro F1 for top-level classification without any deep learning---1.71$\times$ better than the majority-class baseline. This strong baseline separability is reflected in the attention gate of the three-class v4.2/v4.3 model: unlike the binary model where GNN weights dominated at $\sim$60\%, the three-class model settles at CNN $\approx$ 66\% / GNN $\approx$ 34\%. The shift indicates that, once all three TE classes must be discriminated simultaneously, local sequence motifs captured by the CNN become the stronger signal. Within each class, k-mer features perform poorly for superfamily discrimination (macro F1 0.14--0.26 across classes), confirming that the CNN branch drives fine-grained classification.

\subsubsection{Benchmark Exclusion and Final Model (v4.3)}

The final model (v4.3) excludes three species from both training and evaluation: Zebra finch (\textit{Taeniopygia guttata}, \texttt{bTaeGut}), platypus (\textit{Ornithorhynchus anatinus}, \texttt{mOrnAna}), and American alligator (\textit{Alligator mississippiensis}, \texttt{rAllMis}). These species are reserved for an independent external benchmark (Section~\ref{ssec:mini_benchmark}), ensuring that no TEs from these genomes influence learned sequence features. Training was capped at 40 epochs with early stopping on macro F1, and the resulting held-out test set spans 6,141 sequences from the remaining VGP genomes.

\subsubsection{Performance (v4.3)}

The v4.3 model achieves 98.2\% accuracy with a macro F1 of 0.981 on the top-level three-class task (DNA vs LTR vs LINE). All three classes are well-separated: DNA (F1~=~0.978), LTR (F1~=~0.983), and LINE (F1~=~0.983), with per-class accuracy of 97.3\%, 98.2\%, and 98.9\% respectively.

At the superfamily level, accuracy is 89.3\% and macro F1 is 0.839 across 23 superfamilies. LINE subfamilies achieve the highest scores---LINE/CR1 (F1~=~0.992), LINE/RTE (F1~=~0.985), and LINE/L2 (F1~=~0.965)---likely because their distinct compositional profiles make them easily separable. LTR superfamilies show mixed performance: LTR/DIRS (F1~=~0.935), LTR/ERV3 (F1~=~0.936), and LTR/Gypsy (F1~=~0.895) are strong, while the generic \textit{LTR} label (conflating sequences from multiple lineages) is the most challenging at F1~=~0.714. Among DNA transposons, DNA/TcMar-Tc1 (F1~=~0.925) and DNA/hAT (F1~=~0.896) perform well, while DNA/CMC (F1~=~0.494) and LINE/L1-Tx1 (F1~=~0.549) lag behind---consistent with small test support (36 and 69 samples) and biological diversity within those groups.

The most frequent misclassification patterns are biologically interpretable: DNA/hAT sequences are confused with the generic \textit{DNA} label and vice versa ($n$~=~33 and 24); LTR/Gypsy and LTR/Pao are mutually confused ($n$~=~27 and 18); and ERV elements are confused with the unspecified \textit{LTR} label. Crucially, all 15 most common misclassifications occur within the same top-level class (DNA$\rightarrow$DNA, LTR$\rightarrow$LTR, LINE$\rightarrow$LINE), confirming that the model reliably identifies the transposition mechanism even when the precise superfamily is ambiguous.

\subsection{External Benchmark: Mini Dataset Evaluation}\label{ssec:mini_benchmark}

A held-out test set drawn from the same VGP distribution provides evidence of generalisation within the training distribution. A more stringent assessment requires species never encountered in any stage of training. The three species withheld from v4.3---Zebra finch (\textit{T.~guttata}), platypus (\textit{O.~anatinus}), and American alligator (\textit{A.~mississippiensis})---form a mini benchmark spanning the three main amniote lineages (Aves, Mammalia, Reptilia), each with a distinct TE landscape (Table~\ref{tab:mini_benchmark_species}).

\subsubsection{Benchmark Species}

Birds typically have compact genomes dominated by CR1 LINEs with relatively few DNA transposons; the Zebra finch ($\sim$1.0 Gb genome) is a well-studied avian model with an established curated TE library of 543 entries. Platypus is an early-branching monotreme mammal with an unusually rich and active TE repertoire, including families rare in therian mammals; its curated library contains 380 entries. American alligator represents Crocodylia with a moderately sized genome ($\sim$2.2 Gb) and a TE composition intermediate between birds and mammals, with 1,196 curated reference entries providing dense ground-truth coverage.

\begin{table}[h]
\centering
\begin{tabular}{llcc}
\hline
Species & Lineage & Pantera consensi & Curated library \\
\hline
\textit{T. guttata} (Zebra finch) & Aves & 161 & 543 \\
\textit{O. anatinus} (Platypus) & Mammalia & 25 & 380 \\
\textit{A. mississippiensis} (Alligator) & Reptilia & 296 & 1196 \\
\hline
\end{tabular}
\caption{Mini benchmark species, Pantera-derived consensus sequence counts, and curated library sizes.}
\label{tab:mini_benchmark_species}
\end{table}

\subsubsection{Evaluation Protocol}

Evaluation is performed directly on the three held-out curated libraries (\texttt{taeniopygia\_guttata\_lib.fa}, \texttt{platypus\_curated\_lib.fa}, and \texttt{Aligator\_mississippiensis\_lib.fa}) using the best v4.3 checkpoint (epoch 40). Each FASTA header follows RepeatMasker-style annotation (\texttt{Name\#CLASS/SUPERFAMILY}), which provides ground-truth top-level class and superfamily labels. Only sequences whose top-level class is in \{DNA, LTR, LINE\} are evaluated.

For superfamily-level scoring, labels are mapped to the model vocabulary using a hierarchical strategy: exact superfamily match when available, parent-level fallback (e.g., \textit{DNA}, \textit{LTR}, \textit{LINE}) when a specific subtype is unseen, and exclusion from superfamily scoring if no valid mapping exists. Sequences with labels outside the model scope (e.g., tRNA, SINE, RC/Helitron, Unknown) are excluded from the evaluable set.

\subsubsection{Results}

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\hline
Species & Matched sequences & Class accuracy & Superfamily accuracy \\
\hline
\textit{T. guttata} (Zebra finch) & 415 & 33.3\% & 18.8\% \\
\textit{O. anatinus} (Platypus) & 206 & 73.8\% & 41.3\% \\
\textit{A. mississippiensis} (Alligator) & 1003 & 60.1\% & 22.7\% \\
\hline
Overall & 1624 & 55.0\% & 24.1\% \\
\hline
\end{tabular}
\caption{v4.3 classification performance on held-out curated libraries from the mini benchmark. ``Matched sequences'' denotes the evaluable subset after class filtering and label mapping to the model vocabulary.}
\label{tab:mini_benchmark}
\end{table}

Generalization is heterogeneous across species. The model transfers best to platypus at both class and superfamily levels, while Zebra finch shows marked degradation, particularly for LTR elements. Across all three genomes, DNA class recall remains high (approximately 96--97\% in per-class breakdown), but LTR discrimination is substantially weaker, indicating lineage-dependent domain shift in LTR sequence features that is not fully captured by the current training set.


\subsection{Unsupervised Clustering of TE sequences}
[PLACEHOLDER: Unsupervised Clustering]
%TODO: summarize the results from unsupervised clustering experiments, including the UMAP, the loss being used for clustering, and the evaluation metrics (e.g., silhouette score, cluster purity, etc.)

\subsection{[PLACEHOLDER: Results Section 6]}

% ============ DISCUSSION ============
\newpage
\section{Discussion}

[PLACEHOLDER: Discussion]

% ============ MATERIALS AND METHODS ============
\newpage
\section{Materials and Methods}

\subsection{Data}

Training data consisted of transposable element sequences identified by Pantera \cite{sierra2024pantera} from the Vertebrate Genomes Project (VGP) assemblies. Pantera identifies TE families from pangenome polymorphisms by detecting insertions that are absent in some haplotypes, providing a comprehensive catalogue of active and recently active TEs across vertebrate genomes. The dataset comprises approximately 145,000 sequences. Two annotation sets were used: (1) TIR presence/absence labels derived from structural analysis, and (2) TE class labels where sequences are classified as DNA transposons or other TE classes (LTR, LINE, SINE, etc.) based on FASTA header annotations. Approximately 8\% of sequences are DNA transposons, distributed across seven superfamilies with sufficient samples for training (DNA/hAT, TcMar-Tc1, DNA/PIF-Harbinger, DNA/CMC, DNA/MULE, DNA/Kolobok, DNA/PiggyBac).

The sequences have a wide variety of lengths, with a range from 173 bp to 19,907 bp and median at 3,263 bp. To standardize the input, the sequences were placed at random positions within a fixed-length 40kb canvas to enable position augmentation.

\subsection{Model Architectures}

All models employ a tower-based CNN encoder with reverse-complement (RC) invariance. The encoder processes one-hot encoded DNA sequences (5 channels: A, C, G, T, N) through three parallel convolutional towers with kernel sizes 7, 15, and 21 bp, each producing 128 feature maps. Tower outputs are concatenated (384 channels) and mixed via 1-by-1 convolution to 128 channels. Four subsequent dilated convolutional blocks (dilation rates 1, 2, 4, 8) with kernel size 9 expand the receptive field. Each block uses GELU activation, batch normalization, dropout (0.15), and residual connections. Finally, masked global average pooling produces a 128-dimensional embedding.

In current scheme, RC-invariance is achieved through late fusion. The model processes both the original and reverse-complement sequences through the same CNN encoder, and average the outcome embeddings. This allows the model to learn strand-agnostic features without enforcing strict symmetry at the convolutional layer level. 

For the hybrid model (v4), a parallel k-mer GNN branch captures compositional patterns. For each sequence, 7-mer frequencies are extracted from sliding windows (512 bp, stride 256) and hashed to 2048 dimensions using canonical k-mer representation (taking the lexicographically smaller of forward and reverse-complement codes). A relative positional encoding is appended to each window's feature vector. The windows form a chain graph where adjacent windows are connected by bidirectional edges with self-loops. A 3-layer GraphSAGE-style message-passing GNN (hidden dimension 128) processes this graph, aggregating neighborhood information at each layer. Scatter mean pooling produces a sequence-level embedding from the node representations, which is combined with the CNN embeddings via learned attention gating before classification.

\subsection{Training}

Models were trained using AdamW optimizer with learning rate $10^{-3}$, weight decay $10^{-4}$, and cosine annealing schedule. Two strategies are being used to address class imbalance at different levels: the majority class (``None'') is subsampled to balance the binary classification task (DNA vs non-DNA), and (2) inverse square-root class weighting are applied to handle the remaining imbalance across the seven major superfamilies within the DNA class. Label smoothing was used for multi-class heads. 

Training used batch size 32 (16 for hybrid model), and early stopping with patience based on validation macro F1. Cross-validation with 5 folds was employed to ensure robust performance estimates, with a held-out test set reserved for final evaluation. 

An auxiliary boundary prediction head was included during training, predicting the normalized start position, end position, and length of the sequence within the canvas. This auxiliary task (weighted at 0.1) encourages the model to learn positionally-aware representations, though the predictions themselves showed limited accuracy for absolute boundary localization.

\subsection{Evaluation}

Model performance was evaluated using accuracy, balanced accuracy, macro F1 score, and per-class F1 scores. AUROC and AUPRC were computed for binary tasks.

\subsection{Implementation}

All models were implemented in PyTorch and trained on NVIDIA A100 GPUs via CSD3.

% ============ ACKNOWLEDGEMENTS ============
\newpage
\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

[PLACEHOLDER: Acknowledgements]

% ============ REFERENCES ============
\newpage
\section*{References}
\addcontentsline{toc}{section}{References}

% Single-spaced for references
\begin{thebibliography}{99}

\bibitem{sierra2024pantera}
Sierra P, Durbin R (2024) Identification of transposable element families from pangenome polymorphisms. \textit{Mobile DNA} 15:13. https://doi.org/10.1186/s13100-024-00323-y

\end{thebibliography}

\end{document}